---
title: "Log Transformation Process"
format: html
---
## Motivation

To address issues of **non-constant variance** (heteroscedasticity) and **right-skewed distribution** in the response variable, we apply a **log transformation** to the *Pollution Burden Score*. This often stabilizes variance and makes the data more amenable to linear modeling.

## Log-Transformed Model

We fit a linear regression model using the log-transformed response variable:
```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(dplyr)
library(ggplot2)
library(broom)
library(gt)

data <- read_rds(here::here("dataset/cleaned_dataset.rds"))
model_log <- lm(log(`Pollution Burden Score`) ~ ., data = data)


model_perf <- glance(model_log)  # again replace with your actual model

model_perf %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, df, AIC, BIC) %>%
  gt() %>%
  tab_header(
    title = "Statistical Performance of Log Model"
  ) %>%
  fmt_number(decimals = 4)
```

### Statistical Performance of Log Model

The log-transformed model yielded an R-squared of 0.9823, indicating that over 98% of the variance in Pollution Burden Score is explained by the model. The adjusted R-squared of 0.9802 confirms that this strong performance persists even after accounting for the number of predictors. The F-statistic is large (466.39) with a p-value near zero, confirming the model is statistically significant overall.

### Residual Diagnostics
```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 4

# Show two plots side-by-side with better spacing
par(mfrow = c(1, 2))

# Residuals vs Fitted
plot(model_log$fitted.values, model_log$residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted (Log y)")
abline(h = 0, col = "red")

# Q-Q Plot
qqnorm(model_log$residuals)
qqline(model_log$residuals, col = "red")

# Reset plotting layout afterward
par(mfrow = c(1, 1))
```

Despite this strong performance, residual diagnostics revealed patterns suggesting nonlinearity among predictors, and some predictors exhibited skewness or non-constant variance of their own. This prompted us to apply log transformations to both the response and predictors, resulting in a log-log model. The goal was to further improve linearity, stabilize variance across all variables, and enhance model interpretability in terms of percentage change relationships.

[Back](analysis.html)