---
title: "Variable Selection Process"
format: html
---

We compare backward selection and forward model BIC based on AIC to end up with our full model. For detailed analysis on variable selection [click here](var_select.html)

### Backward Selection
```{r}
#| echo: false
#| message: false
#| warning: false
#| results: "hide"

library(tidyverse)
library(dplyr)
library(ggplot2)
library(broom)
library(gt)

data <- read_rds(here::here("dataset/cleaned_dataset.rds"))

log_data <- data %>%
  mutate(across(
    c(`Pollution Burden Score`,
      Ozone, `PM2.5`, `Diesel PM`, Traffic,
      `Cleanup Sites`, `Haz. Waste`, `Solid Waste`,
      Asthma, `Low Birth Weight`, `Cardiovascular Disease`,
      `Children < 10 years (%)`, `Elderly > 64 years (%)`, `White (%)`,
      `Hispanic (%)`, `African American (%)`, `Asian American (%)`),
    ~ log(. + 1)
  ))

log_log_model <- lm(`Pollution Burden Score` ~ 
  Ozone + `PM2.5` + `Diesel PM` + Traffic +
  `Cleanup Sites` + `Haz. Waste` + `Solid Waste` +
  Asthma + `Low Birth Weight` + `Cardiovascular Disease` +
  `Children < 10 years (%)` + `Elderly > 64 years (%)` + 
  `Hispanic (%)` + `African American (%)` + `Asian American (%)` + `White (%)`,
  `California County`,
  data = log_data)

```


```{r}
#| echo: false
#| message: false
#| warning: false


library(broom)
library(gt)

quiet_step <- function(...) {
  capture.output(result <- step(..., trace = 0))
  return(result)
}

backward_model <- quiet_step(
  step(log_log_model, direction = "backward")
)

glance(backward_model) |>
  select(r.squared, adj.r.squared, sigma, statistic, p.value, df, AIC, BIC) |>
  gt() |>
  tab_header(title = "Statistical Performance: Backward Model") |>
  fmt_number(columns = everything(), decimals = 4)
```

### Forward Model BIC Selection

```{r}
#| echo: false
#| message: false
#| warning: false
library(stats)     # for step()

n <- nrow(log_data)  # get sample size for BIC
k_bic <- log(n)

quiet_step_bic <- function(...) {
  capture.output(result <- step(..., direction = "forward", trace = 0, k = k_bic))
  return(result)
}

forward_model_bic <- quiet_step_bic(log_log_model)

glance(forward_model_bic) |>
  select(r.squared, adj.r.squared, sigma, statistic, p.value, df, AIC, BIC) |>
  gt() |>
  tab_header(title = "Statistical Performance: Forward Model (BIC)") |>
  fmt_number(columns = everything(), decimals = 4)
```

Both the **Forward (BIC)** and **Backward** models yield nearly identical statistical performance, with the same R² (0.8600) and adjusted R² (0.8597), as well as nearly identical residual standard error (σ = 0.0213). However, the **Backward model** shows a **higher F-statistic** (3518 vs. 3078), indicating stronger overall explanatory power.

More importantly, the **Backward model has a slightly lower BIC (-38,943.15 vs. -38,926.68)** and uses **fewer predictors (14 vs. 16)**, which makes it **more parsimonious** while maintaining the same predictive accuracy.

## Conclusion

We select the **Backward model** as the final model. It provides equivalent explanatory performance with fewer predictors and a better BIC, striking an optimal balance between model complexity and goodness of fit.

[Back](analysis.html)